---
title: "clustering"
author: "Anson Wong"
date: '2023-04-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "output/")
```

Load packages
```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(SeuratData))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(scCustomize))
suppressPackageStartupMessages(library(clustree))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(Polychrome))
suppressPackageStartupMessages(library(pals))
suppressPackageStartupMessages(library(BiocStyle))
suppressPackageStartupMessages(library(glue))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(scuttle))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(EnsDb.Hsapiens.v86))
suppressPackageStartupMessages(library(ensembldb))
suppressPackageStartupMessages(library(Homo.sapiens))
suppressPackageStartupMessages(library(msigdbr))
suppressPackageStartupMessages(library(EGSEA))
suppressPackageStartupMessages(library(speckle))
suppressPackageStartupMessages(library(here))
```

Read objects
```{r}
# Read object
sce <- readRDS((here("data_230503","SCEs","G000323.preprocessed.SCE.rds")))
```

Identify uninformative genes
```{r}
# Some useful gene sets
mito_set <- rownames(sce)[which(rowData(sce)$CHR == "MT")]
ribo_set <- grep("^RP(S|L)", rownames(sce), value = TRUE)
# NOTE: A more curated approach for identifying ribosomal protein genes
#       (https://github.com/Bioconductor/OrchestratingSingleCellAnalysis-base/blob/ae201bf26e3e4fa82d9165d8abf4f4dc4b8e5a68/feature-selection.Rmd#L376-L380)
c2_sets <- msigdbr(species = "Homo sapiens", category = "C2")
ribo_set <- union(
  ribo_set,
  c2_sets[c2_sets$gs_name == "KEGG_RIBOSOME", ]$human_gene_symbol)
sex_set <- rownames(sce)[any(rowData(sce)$ENSEMBL.SEQNAME %in% c("X", "Y"))]
pseudogene_set <- rownames(sce)[
  any(grepl("pseudogene", rowData(sce)$ENSEMBL.GENEBIOTYPE))]

```

Calculate quality control metrics
```{r}
is_mito <- rownames(sce) %in% mito_set
is_ribo <- rownames(sce) %in% ribo_set
head(colData(sce)) %>% knitr::kable()

```

Discard uninformative genes
```{r}
uninformative <- is_mito | is_ribo | rownames(sce) %in% sex_set | rownames(sce) %in% pseudogene_set
# # original:
# junk <- sce$HTO_demuxmix %in% c("multiplet", "negative") | (sce$cell_status == "empty_droplet")
# sceFlt <- sce[!uninformative, !junk] 
junk <- sce$HTO_HTODemux %in% c("Doublet", "Negative")
sceFlt <- sce[!uninformative, !junk]

```

Remove low-abundance genes
```{r}
numCells <- nexprs(sceFlt, byrow = TRUE)
keep <- numCells > 20
#sum(keep)
sceFlt <- sceFlt[keep,]
```

Convert to Seurat object
```{r}
counts <- counts(sceFlt)
rownames(counts) <- rowData(sceFlt)$Symbol
seu <- CreateSeuratObject(counts = counts, meta.data = data.frame(colData(sceFlt)))
dim(seu)
```

Normalize
```{r}
DefaultAssay(seu) <- "RNA"
seu <- SCTransform(seu,
                   vst.flavor="v2",
                   variable.features.n = 3000,
                   vars.to.regress = c("subsets_Mito_percent", "subsets_Ribo_percent"))

DimPlot(seu, reduction = "umap", group.by = "orig.ident",
        label = FALSE)
```

Linear dimensional reduction
```{r}
p1 <- DimPlot(seu, reduction = "pca", group.by = "orig.ident")
p2 <- DimPlot(seu, reduction = "pca", dims = c(1,3), group.by = "orig.ident")
p3 <- DimPlot(seu, reduction = "pca", dims = c(2,3), group.by = "orig.ident")
p4 <- DimPlot(seu, reduction = "pca", dims = c(3,4), group.by = "orig.ident")

((p1 | p2) / (p3 | p4)) + plot_layout(guides = "collect") &
  theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 8))

DimHeatmap(seu, dims = 1:20, cells = 500, balanced = TRUE)

ElbowPlot(seu, ndims = 30)
```

Cluster the cells
```{r}
out <- here("data/SCEs/G000323.clustered.SEU.rds")
seu <- FindNeighbors(seu, reduction = "pca", dims = 1:30)
plan("multicore",workers=20)
options(future.seed=TRUE)
seu <- FindClusters(seu, algorithm = 4,  method = "igraph", 
                    resolution = seq(0.1, 2, by = 0.1))
ggsave(plot=clustree(seu, prefix="SCT_snn_res."), device="pdf",
       width = 10,
       height = 18,
       file="all.clustree.alg4.230425.pdf")

```

Non-linear dimensional reduction
```{r}
seu <- RunUMAP(seu, dims = 1:30)

# Visualize UMAP
colours <- c(brewer.paired(12), brewer.dark2(8), brewer.set2(8), rev(brewer.accent(8)))
p1 <- DimPlot(seu, reduction = "umap",
              cols = colours,
              group.by = "SCT_snn_res.0.9", label = TRUE,
              label.size = 2.5, repel = TRUE) +
  NoLegend() +
  ggtitle("SCT_snn_res.0.9")  +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        title = element_text(size = 9))
p2 <- DimPlot(seu, reduction = "umap",
              cols = colours,
              group.by = "SCT_snn_res.1", label = TRUE,
              label.box = TRUE,
              label.size = 2.5, repel = TRUE) +
  NoLegend() +
  ggtitle("SCT_snn_res.1")  +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        title = element_text(size = 9))
ggsave(plot=p1 + p2, device="pdf", height=6, width=12,
       file="all.umap.alg4.res0.9_1.230425.pdf")

if(!file.exists(out)) {
  saveRDS(seu, file = out)
}

```

Annotate data using Zilionis's reference
```{r}
zilSct <- readRDS("/group/canc2/anson/reference/annotation/GSE127465_RAW/ZILIONIS.sct_normalised.SEU.rds")
options(future.globals.maxSize = 5000000000, future.seed=TRUE)
anchors <- FindTransferAnchors(reference = zilSct, query = seu,
                               dims = 1:30, reference.reduction = "pca",
                               normalization.method = "SCT")
seu <- MapQuery(anchorset = anchors, reference = zilSct,
                query = seu,
                refdata = list(celltype = "Major.cell.type"),
                reference.reduction = "pca",
                reduction.model = "umap")
out <- here('data/SCEs/G000323.clustered.Zilionis.SEU.rds')
if(!file.exists(out)) {
  saveRDS(seu, out)
}

# Visualize
p1 <- DimPlot(seu, reduction = 'umap',
        label = TRUE, repel = TRUE,
        label.size = 2.5,
        group.by = "predicted.celltype") + NoLegend()
ggsave(plot=p1, device="pdf", height=6, width=12,
       file="all.umap.alg4.Zilionis.230425.pdf")

```

Annotate data using HLCA v2 [requires R v4.2.2]
```{r}
# load libraries
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(Azimuth))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(forcats))

# read object
seu <- readRDS(here('data','SCEs','G000323.clustered.Zilionis.SEU.rds'))
# Slim down object
DefaultAssay(seu) <- "RNA"
seu.diet <- DietSeurat(seu, assays = "RNA")
# run Azimuth
seu.diet <- RunAzimuth(seu.diet, 
                       "/group/canc2/anson/reference/annotation/HLCA_v2", 
                       umap.name = 'azimuth.umap',
                       verbose = TRUE, 
                       assay = "RNA", 
                       k.weight = 50, 
                       n.trees = 20, 
                       mapping.score.k = 100)
# rds format from the web Azimuth app:
# list(umap=umap, predicted.df=df)
umap <- seu.diet@reductions$azimuth.umap
df <- data.frame(row.names=rownames(seu.diet@meta.data),
                 cell = rownames(seu.diet@meta.data), 
                 predicted.ann_level_1=seu.diet$predicted.ann_level_1,
                 predicted.ann_level_2=seu.diet$predicted.ann_level_2,
                 predicted.ann_level_3=seu.diet$predicted.ann_level_3,
                 predicted.ann_level_4=seu.diet$predicted.ann_level_4,
                 predicted.ann_level_5=seu.diet$predicted.ann_level_5,
                 predicted.ann_finest_level=seu.diet$predicted.ann_finest_level,
                 predicted.ann_level_1.score=seu.diet$predicted.ann_level_1.score,
                 predicted.ann_level_2.score=seu.diet$predicted.ann_level_2.score,
                 predicted.ann_level_3.score=seu.diet$predicted.ann_level_3.score,
                 predicted.ann_level_4.score=seu.diet$predicted.ann_level_4.score,
                 predicted.ann_finest_level.score=seu.diet$predicted.ann_finest_level.score,
                 mapping.score=seu.diet$mapping.score)

# generate the azimuth result (which is a list)
azimuth.result <- list(umap = umap, pred.df = df)

# save the list
saveRDS(azimuth.result, file = here('data','SCEs','azimuth.result.rds'))

# add annotation
seu <- AddAzimuthResults(seu, filename=here('data','SCEs','azimuth.result.rds'))
seu$predicted.ann_level_1 <- fct_drop(seu$predicted.ann_level_1)
seu$predicted.ann_level_2 <- fct_drop(seu$predicted.ann_level_2)
seu$predicted.ann_level_3 <- fct_drop(seu$predicted.ann_level_3)
seu$predicted.ann_level_4 <- fct_drop(seu$predicted.ann_level_4)
seu$predicted.ann_finest_level <- fct_drop(seu$predicted.ann_finest_level)

out <- here('data','SCEs','G000323.clustered.Zilionis.HLCA.SEU.rds')

# save object
if(!file.exists(out)) {
  saveRDS(seu, out)
}

```

Split and save object
```{r}
seu <- readRDS(here('data','SCEs','G000323.clustered.Zilionis.HLCA.SEU.rds'))
labels <- levels(factor(seu$predicted.ann_level_3))
macrophages <- c("Macrophages")
tcells <- c("T cell lineage", "Innate lymphoid cell NK")
#myeloid <- c("Dendritic cells", "Mast cells", "Monocytes", "Macrophages")
#epithelial <- c("Basal", "Multiciliated lineage", "Rare", "Secretory")

# lung <- c("AT1", "Rare", "Secretory", "Basal", "EC capillary", #old
#           "EC venous", "Multiciliated lineage")

subList <- list(macrophages = macrophages,
                tcells = tcells, 
                others = labels[!labels %in% c(macrophages, tcells)]
                )

Idents(seu) <- "predicted.ann_level_3"

for(sub in names(subList)){
  out <- here(glue("data/SCEs/G000323.clustered.Zilionis.HLCA.{sub}.SEU.rds"))
  message(sub)
  a <- subset(seu, idents = subList[[sub]])
  DefaultAssay(a) <- "RNA"
  if(!file.exists(out)){
    saveRDS(DietSeurat(a, assays = "RNA", dimreducs = NULL, graphs = NULL), out)
  }
}

```

Visualize UMAP
```{r}
colours <- c(brewer.paired(12), brewer.dark2(8), brewer.set2(8), rev(brewer.accent(8)))
# Visualize annotation
p1 <- DimPlot(seu, reduction = "umap",
              cols = colours,
              group.by = "predicted.ann_level_3", label = FALSE,
              label.size = 2.5, repel = TRUE, label.box = FALSE) +
  #NoLegend() +
  ggtitle("predicted.ann_level_3")  +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        title = element_text(size = 9))
p2 <- DimPlot(seu, reduction = "umap",
              cols = colours,
              group.by = "predicted.ann_level_4", label = FALSE,
              label.size = 2.5, repel = TRUE, label.box = FALSE) +
  #NoLegend() +
  ggtitle("predicted.ann_level_4")  +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        title = element_text(size = 9))
p3 <- DimPlot(seu, reduction = "umap",
              cols = colours,
              group.by = "predicted.ann_finest_level", label = FALSE,
              label.size = 2.5, repel = TRUE, label.box = FALSE) +
  #NoLegend() +
  ggtitle("predicted.ann_finest_level")  +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        title = element_text(size = 9))

p4 <- DimPlot(seu, reduction = "umap",
              cols = colours,
              group.by = "SCT_snn_res.0.9", label = TRUE,
              label.size = 2.5, repel = TRUE, label.box = TRUE) +
  NoLegend() +
  ggtitle("SCT_snn_res.0.9")  +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        title = element_text(size = 9))
p2/p3 + p4
#+ NoLegend()
ggsave(plot=p2/p3 + p4, file="mac.umap.jpeg", device="jpeg", dpi=300, height=12, width=6)


factor(seu$predicted.celltype)

table(seu$predicted.celltype) %>% knitr::kable()

```












